---
title: >
  The `KEGGemUP` package: building KEGG pathway graphs and mapping real data results
author:
  - name: Edoardo Filippi
    affiliation: 
    - Institute of Medical Biostatistics, Epidemiology and Informatics (IMBEI), Mainz
    - Deparment of Nephrology, Mainz University Medical Center
    email: edoardo.filippi@uni-mainz.de
  - name: Federico Marini
    affiliation: 
    - Institute of Medical Biostatistics, Epidemiology and Informatics (IMBEI), Mainz
    email: federico.marini@uni-mainz.de
date: "`r BiocStyle::doc_date()`"
package: "`r BiocStyle::pkg_ver('KEGGemUP')`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{kegg-graph}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

**Compiled date**: `r Sys.Date()`

**Last edited**: 15-12-2025

**License**: `r packageDescription("KEGGemUP")[["License"]]`

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Installation

The first step is installing the package.
If you dont have the package installed you can do so by running:

```{r install-package, eval=FALSE}
install.packages("remotes")     
remotes::install_github("edo98811/KEGGemUP")
```

Then load it.

```{r}
library("KEGGemUP")
```

KEGG pathways are used in many bioinformatics contexts, this package package can make it very easy to implement these in the analysis of real data.

A very important part of bioinformatics analyses is both data integration and visiazionation. KEGGemUP aims to facilitate these tasks by providing functions to parse KEGG pathways and build graph objects, the idea is then to use these graph to map on them thre results of differential expression analyses. 

KEGG database uses pathway ids to identify each pathway, for example "hsa00563" is the KEGG pathway ID for "Glycosylphosphatidylinositol (GPI)-anchor biosynthesis" in humans. You can use these pathway IDs to download the KGML files for each pathway and then parse them to build graph objects. The id must have the organism prefix, for example "hsa" for human pathways, "mmu" for mouse pathways, etc.

These vignette assumes you have already performed a differential expression analysis and have the results available as a `data.frame` (which you can easily obtain from edgeR or limma by running `as.data.frame()` on the table of the results for each object). To reproduce this situation we will now build an example of a differential expression result using the macrophage dataset from the macrophage package: [bioconductor page](https://www.bioconductor.org/packages/release/data/experiment/html/macrophage.html)

To learn more about graph manipulation you can refer to the igraph and visNetwork documentation: 

- [igraph](https://r.igraph.org) Is a pacakge for creating and manipulating graphs in R.
- [visNetwork](https://datastorm-open.github.io/visNetwork/) Is a package for interactive visualization of graphs in R, an R interface to the javascript library vis.js.


```{r load-example-results}
message("--- Loading packages...")

suppressPackageStartupMessages({
  library("macrophage")
  library("org.Hs.eg.db")
  library("SummarizedExperiment")
  library("AnnotationDbi")
  library("clusterProfiler")
  library("limma")
  library("edgeR")
})
message("- Done!")

# Load the macrophage dataset ---------------------------------------------------
data(gse)
rownames(gse) <- substr(rownames(gse), 1, 15)  # truncate rownames at 15 characters

# limma analysis ---------------------------------------------------------------
condition <- factor(colData(gse)[, "condition_name"])
design <- model.matrix(~0 + condition)

contrast.matrix <- makeContrasts(
  IFNg_vs_naive = conditionIFNg - conditionnaive,
  levels = design
)

dge <- DGEList(assay(gse))
dge <- calcNormFactors(dge)

cutoff <- 1
drop <- which(apply(cpm(dge), 1, max) < cutoff)
dge <- dge[-drop, ]

voom_mat <- voom(dge, design, plot = FALSE)
fit <- lmFit(voom_mat, design)
fit2 <- contrasts.fit(fit, contrast.matrix)
fit2 <- eBayes(fit2)  # Empirical Bayes moderation

# Gene annotation ---------------------------------------------------------------

anns <- AnnotationDbi::select(
  org.Hs.eg.db,
  keys = rownames(gse),
  columns = c("SYMBOL", "ENTREZID"),
  keytype = "ENSEMBL",
  multiVals = "first"
)

# limma results ---------------------------------------------------------------
res_macrophage_IFNg_vs_naive_limma <- topTable(
  fit2,
  coef = "IFNg_vs_naive",
  adjust = "fdr",
  number = Inf,
  confint = TRUE
)

res_macrophage_IFNg_vs_naive_limma$ENTREZID <- anns$ENTREZID[match(rownames(res_macrophage_IFNg_vs_naive_limma), anns$ENSEMBL)]

# Enrichment analysis ----------------------------------------------------------
de_entrez_IFNg_vs_naive_genes <- anns $ENTREZID[
  (!is.na(res_macrophage_IFNg_vs_naive_limma$adj.P.Val)) &
    (res_macrophage_IFNg_vs_naive_limma$adj.P.Valj <= 0.05)
]

```


## Functions to parse KGML files

KGML is the format that KEGG uses to save the pathway structure and it is what this package interfaces itself with. There is an exported function to download KGML files from KEGG, `download_kgml()`, which you can use to get the KGML file for a given pathway ID. You can pass to the function the KEGG pathway ID and the directory where to save the file.

As an example we will download the KGML file for the KEGG pathway "Glycosylphosphatidylinositol (GPI)-anchor biosynthesis" in humans, which has the KEGG pathway ID "hsa00563".

```{r download-kgml}
kgml_file <- download_kgml("hsa00563", directory = tempdir())  # KEGG pathway ID for "Glycosylphosphatidylinositol (GPI)-anchor biosynthesis"
kgml_file
``` 

You can use these functions to parse KGML files directly. From these you can build a graph object if you wish to do so and you have expertese with graph analysis in R. The first function is `parse_kgml_entries()` which parses the nodes of the pathway graph from a KGML file. The second function is `parse_kgml_relations()` which parses the edges of the pathway graph from a KGML file.


```{r kgml-parsing}
nodes_df <- parse_kgml_entries(kgml_file)
edges_df <- parse_kgml_relations(kgml_file)
```


### The output data.frame frame for nodes

Here you can see the first 5 columns of the dataframe that you get by parsing the nodes from a KGML file. It is a `data.frame` where each row represents a node in the KEGG pathway graph. No operations are done on the data, it is just a direct parsing of the KGML file.

```{r}
knitr::kable(head(nodes_df))
```


### The output data.frame frame for edges

Here you can see the first 5 columns of the dataframe that you get by parsing the edges from a KGML file. It is a `data.frame` where each row represents an edge in the KEGG pathway graph. No operations are done on the data, it is just parsing of the KGML file while keeping all the information from it.

```{r}
knitr::kable(head(edges_df))
```

## Build a graph from a pathway ID and map results to nodes

To map the differential expression results to the nodes of a KEGG pathway graph you can use `map_results_to_nodes()`. The input of this function is a graph object built with `kegg_to_graph()` and a list of differential expression results tables or a single differential expression results table.

You can also pass as input to `map_results_to_nodes()` a single `data.frame` with the differential expression results. This dataframe must contain at least two columns: one with the KEGG feature IDs (without organism prefix) and another with the values to map to the nodes (e.g., log2 fold changes). You can pass to the function the names of these columns if they differ from the default ones with the parameters `feature_column` and `value_column`.

Note that the KEGG IDs without organism prefix are the the ENTREZ IDs for genes. For other feature types (e.g., compounds) you will need to make sure that the IDs in your differential expression results table match the KEGG IDs used in the graph.

The defualt parameters are:

- `feature_column`: "KEGG_ids"
- `value_column`: "log2FoldChange"

If you have multiple differential expression results tables (for example if you have one metabolomics and one transcriptomics) to map to the nodes you can pass a list of lists. Each sublist must contain the following elements:

- `de_table`: a data.frame with the differential expression results.
- `value_column`: the name of the column in de_table containing the values to map to the nodes.
- `feature_column`: the name of the column in de_table containing the feature IDs (e.g., ENTREZ IDs) that correspond to the KEGG ids in the graph (without organism prefix).

This is an example of how to build such a list of differential expression results tables. In this case there is only one element, but ideally you would have one for each omics layer or source you want to map to the graph nodes.

```{r}
de_results_list <-list(
  trans_limma = list(
    de_table = data.frame(res_macrophage_IFNg_vs_naive_limma),
    value_column = "logFC",
    feature_column = "ENTREZID"
  )
)
```


### Example of usage with the list of DE results tables

We will now take a KEGG pathway from the enrichment results we built earlier and map the differential expression results to its nodes. As you can see you simply need to pass to the function the graph object and the list of differential expression results tables that we defined before.

The output of the first fucntion should be an igraph object, which is what we specify with the parameter `return_type = "igraph"`. It can also be a visNetwork object if you set `return_type = "visNetwork"`. The visNetwork object is useful for interactive visualization of the graph, but to map the results to the nodes we need to start from an igraph object.
The output of the second function will be a visNetwork object with the differential expression results mapped to the nodes. In this case we set `return_type = "visNetwork"` to get a visNetwork object as output, which then we can visualize directly.

```{r use-package}
pathway <- "hsa00563"  
graph <- kegg_to_graph(pathway, return_type = "igraph")
graph_visnetwork <- map_results_to_graph(graph, de_results_list, return_type = "visNetwork")
graph_visnetwork
```

We could also have the first function return a visNetwork object directly by setting `return_type = "visNetwork"` in `kegg_to_graph()`. In this case we can immediately visualize the graph but to perfom the mapping of the differential expression results to the nodes we need an igraph object as input to `map_results_to_graph()`.

### Example of usage with a single DE results table

Let's first build a filtered differential expression results table with only the significant results.

```{r}
de_results_limma <-  data.frame(res_macrophage_IFNg_vs_naive_limma)
```

Then we will call the function with this table as input. The parameter `feature_column` and `value_column` are set to match the column names in our differential expression results table.

```{r use-package-2}
graph <- kegg_to_graph(pathway, return_type = "igraph")
graph_visnetwork <- map_results_to_graph(graph, de_results_limma, feature_column = "ENTREZID", value_column = "logFC", return_type = "visNetwork")
graph_visnetwork
```

You can also control the palette that is used to map the values to colors on the nodes with the parameter `palette`. The default is "RdBu" from RColorBrewer, but you can use any palette supported by RColorBrewer.
To see the available palettes you can run `RColorBrewer::display.brewer.all()`.
You can also visit this page: [RColorBrewer palettes](https://r-graph-gallery.com/38-rcolorbrewers-palettes.html).

```{r use-package-3}
graph <- kegg_to_graph(pathway, return_type = "igraph")
graph_visnetwork <- map_results_to_graph(graph, de_results_limma, feature_column = "ENTREZID", value_column = "logFC", return_type = "visNetwork", palette = "PiYG")
graph_visnetwork
```

## Session info
```{r}
sessionInfo()
```